---
title: "Programming Exercise 1: Linear Regression"
author: ""
date: "Sep 2, 2020"
output: 
    html_document:
        toc: true
        mathjax: local
        self_contained: FALSE
---

```{r opt, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache=TRUE, fig.align="center")
sources <- c("computeCost-solution.r","gradientDescent-solution.r",
             "plotData-solution.r","warmUpExercise-solution.r")
sources <- c(sources, "featureNormalize-solution.r", "computeCostMulti-solution.r",
	"gradientDescentMulti-solution.r", "normalEqn-solution.r", "../lib/pinv.r")
invisible(lapply(sources, source))
```

This programming exercise instruction was originally developed and written by Prof. Andrew
Ng as part of his machine learning [course](https://www.coursera.org/learn/machine-learning) on Coursera platform. I have adapted
the instruction for R language, so that its users, including myself, could also take and benefit from the course.

## Introduction
In this exercise, you will implement linear regression and get to see it work
on data. Before starting on this programming exercise, we strongly recommend
watching the video lectures and completing the review questions for the
associated topics.
To get started with the exercise, you will need to download the starter
code and unzip its contents to the directory where you wish to complete the
exercise. If needed, use the `setwd()` command in R to change to
this directory before starting this exercise.

Files included in this exercise:

- `ex1.r` - R script that steps you through the exercise
- `ex1_multi.r` - R script for the later parts of the exercise
- `ex1data1.txt` - Dataset for linear regression with one variable
- `ex1data2.txt` - Dataset for linear regression with multiple variables
- `submit.r` - Submission script that sends your solutions to our servers
- [⋆] `warmUpExercise.r` - Simple example function in R
- [⋆] `plotData.r` - Function to display the dataset
- [⋆] `computeCost.r` - Function to compute the cost of linear regression
- [⋆] `gradientDescent.r` - Function to run gradient descent
- [†] `computeCostMulti.r` - Cost function for multiple variables
- [†] `gradientDescentMulti.r` - Gradient descent for multiple variables
- [†] `featureNormalize.r` - Function to normalize features
- [†] `normalEqn.r` - Function to compute the normal equations

⋆ indicates files you will need to complete and † indicates optional exercises

Throughout the exercise, you will be using the scripts `ex1.r` and `ex1_multi.r`.
These scripts set up the dataset for the problems and make calls to functions
that you will write. You do not need to modify either of them. You are only
required to modify functions in other files, by following the instructions in
this assignment.
For this programming exercise, you are only required to complete the first
part of the exercise to implement linear regression with one variable. The
second part of the exercise, which is optional, covers linear regression with
multiple variables.

### Where to get help
The exercises in this course use R, a high-level programming language
well-suited for numerical computations. If you do not have R installed, you may
download a Windows installer from [R-project](https://cran.r-project.org/bin/windows/base/) website. [R-Studio](https://rstudio.com/products/rstudio/download/) is a free and open-source R integrated development environment (IDE) making R script development a bit easier when compared to R basic GUI. You may start from the `.rproj` (a R-Studio project file) in each exercise directory. 
At the R command line, typing `help` followed by a
function name displays documentation for that function. For example,
`help('plot')` or simply `?plot` will bring up help information for plotting.
Further documentation for R functions can be found at the R documentation
pages. 

## 1 Simple R function

The first part of `ex1.r` gives you practice with R syntax and
the homework submission process. In the file `warmUpExercise.r`, you will
find the outline of an R function. Modify it to return a $5 \times 5$ identity matrix by filling in the following code:

```{r eye, eval=F}
A <- diag(5)
```

When you are finished, run `ex1.r` (assuming you are in the correct directory, type `source("ex1.r")` at the R prompt) and you should see output similar to the following:

```{r, echo=F}
diag(5)
```

Now `ex1.r` will pause until you press any key, and then will run the code
for the next part of the assignment. If you wish to quit, typing <kbd>ctrl-c</kbd> will
stop the program in the middle of its run.

### 1.1 Submitting Solutions

After completing a part of the exercise, you can submit your solutions for
grading by typing submit at the R command line. The submission script will prompt you for your login e-mail and submission token
and ask you which files you want to submit. You can obtain a submission
token from the web page for the assignment.

*You should now submit your solutions.*

You are allowed to submit your solutions multiple times, and we will take
only the highest score into consideration.

## 2 Linear regression with one variable

In this part of this exercise, you will implement linear regression with one
variable to predict profits for a food truck. Suppose you are the CEO of a
restaurant franchise and are considering different cities for opening a new
outlet. The chain already has trucks in various cities and you have data for
profits and populations from the cities.

You would like to use this data to help you select which city to expand to next.
The file `ex1data1.txt` contains the dataset for our linear regression problem. The first column is the population of a city and the second column is
the profit of a food truck in that city. A negative value for profit indicates a loss.
The `ex1.r` script has already been set up to load this data for you.

### 2.1 Plotting the Data

Before starting on any task, it is often useful to understand the data by
visualizing it. For this dataset, you can use a scatter plot to visualize the
data, since it has only two properties to plot (profit and population). (Many
other problems that you will encounter in real life are multi-dimensional and
can't be plotted on a 2-d plot.)
In `ex1.r`, the dataset is loaded from the data file into the variables X
and y:

```{r}
data <- read.table("ex1data1.txt",sep=',')
X <- data[, 1]
y <- data[, 2]
m <- length(y) # number of training examples
```

Next, the script calls the plotData function to create a scatter plot of
the data. Your job is to complete `plotData.r` to draw the plot; modify the
file and fill in the following code:

```{r, eval=FALSE}
plot(x, y, col = "red", pch = 4, cex = 1.1, lwd = 2,
    xlab = 'Profit in $10,000s',
    ylab = 'Population of City in 10,000s')
```

Now, when you continue to run `ex1.r`, our end result should look like
Figure 1, with the same red "x" markers and axis labels.
To learn more about the plot command, you can type `?plot` at the
R command prompt or to search for plotting documentation. (To change the markers to red "x", we used the option  `pch=4` 
together with the plot command, i.e., `plot(..,[your options here],.., pch="4")` )

```{r, echo=FALSE, fig.cap="Figure 1: Scatter plot of training data"}
par(mar=c(3,3,.5,0), mgp=c(1.5,.5,0))
plotData(X, y)
```

### 2.2 Gradient Descent

In this part, you will fit the linear regression parameters $\theta$ to our dataset using gradient descent.

#### 2.2.1 Update Equations

The objective of linear regression is to minimize the cost function

$$ J(\theta)= \frac{1}{2m} \sum_{i=1}^m (h_\theta (x^{(i)})-y^{(i)})^2 $$

where the hypothesis $h_\theta(x)$ is given by the linear model

$$ h_\theta(x)=\theta^Tx=\theta_0+\theta_1x_1 $$

Recall that the parameters of your model are the $\theta_j$ values. These are
the values you will adjust to minimize cost $J(\theta)$. One way to do this is to
use the batch gradient descent algorithm. In batch gradient descent, each
iteration performs the update:

<center>

$$ \theta_j:=\theta_j-\alpha\frac{1}{m}\sum_{i=1}^m (h_\theta (x^{(i)})-y^{(i)})x_j^{(i)}\qquad\text{(simultaneously update $\theta_j$ for all $j$).} $$

</center>

With each step of gradient descent, your parameters $\theta_j$ come closer to the optimal values that will achieve the lowest cost $J(\theta)$.

___
**Implementation Note:** We store each example as a row in the the X
matrix in R. To take into account the intercept term ($\theta_0$),
we add an additional first column to X and set it to all ones. This allows
us to treat $\theta_0$ as simply another 'feature'.

___

#### 2.2.2 Implementation

In `ex1.r`, we have already set up the data for linear regression.
In the following lines, we add another dimension to our data to accommodate the
$\theta_0$ intercept term. We also initialize the initial parameters to 0 and the
learning rate alpha to 0.02.

```{r}
X <- cbind(rep(1,m),X) # Add a column of ones to x
X <- as.matrix(X)
# initialize fitting parameters
theta <- c(8,3)

# Some gradient descent settings
iterations <- 1500
alpha <- 0.02
```

#### 2.2.3 Computing the cost $J(\theta)$

As you perform gradient descent to learn minimize the cost function $J(\theta)$,
it is helpful to monitor the convergence by computing the cost.
In this section, you will implement a function to calculate $J(\theta)$ so you can check the
convergence of your gradient descent implementation.
Your next task is to complete the code in the file `computeCost.r`, which
is a function that computes $J(\theta)$. As you are doing this, remember that the
variables X and y are not scalar values, but matrices whose rows represent
the examples from the training set.
Once you have completed the function, the next step in `ex1.r` will run
`computeCost` once using $\theta$ initialized to zeros, and you will see the cost
printed to the screen.
You should expect to see a cost of 32.07.

*You should now submit your solutions.*

#### 2.2.4 Gradient descent

Next, you will implement gradient descent in the file `gradientDescent.r`.
The loop structure has been written for you, and you only need to supply
the updates to $\theta$ within each iteration.
As you program, make sure you understand what you are trying to optimize and what is being updated. Keep in mind that the cost $J(\theta)$ is parameterized by the vector $\theta$, not X and y. That is, we minimize the value of $J(\theta)$
by changing the values of the vector $\theta$, not by changing X or y. Refer to the
equations in this handout and to the video lectures if you are uncertain.
A good way to verify that gradient descent is working correctly is to look
at the value of $J(\theta)$ and check that it is decreasing with each step. The
starter code for `gradientDescent.r` calls `computeCost` on every iteration
and prints the cost. Assuming you have implemented gradient descent and
`computeCost` correctly, your value of $J(\theta)$ should never increase, and should
converge to a steady value by the end of the algorithm.
After you are finished, `ex1.r` will use your final parameters to plot the
linear fit. The result should look something like Figure 2.
Your final values for $\theta$ will also be used to make predictions on profits in
areas of 35,000 and 70,000 people. Note the way that the following lines in
`ex1.r` uses matrix multiplication, rather than explicit summation or looping, to calculate the predictions. This is an example of code vectorization in
R.

*You should now submit your solutions.*

```{r}
predict1 <- c(1, 3.5) %*% theta
predict2 <- c(1, 7) %*% theta
```

### 2.3 Debugging

Here are some things to keep in mind as you implement gradient descent:

- R vector indices start from one, not zero. If you're storing $\theta_0$ and $\theta_1$ in a vector called `theta`, the values will be `theta[1]` and `theta[2]`.

- If you are seeing many errors at runtime, inspect your matrix operations
to make sure that you're adding and multiplying matrices of compatible dimensions.
Printing the dimensions of matrices with the `dim()` command and the length of vectors with `length()` command will help you debug.

```{r , echo=FALSE, fig.cap="Figure 2: Training data with linear regression fit"}
iterations <- 1500
alpha <- 0.02
cost <- computeCost(X, y, theta)
gd <- gradientDescent(X, y, theta, alpha, iterations)
theta <- gd$theta
J_history <- gd$J_history
theta_history <- gd$theta_history

par(mar=c(3,3,.5,0), mgp=c(1.5,.5,0))
plotData(X[,-1], y)
lines(X[, 2], X  %*% theta, col="blue")
legend("bottomright", c('Training data', 'Linear regression'), pch=c(4,NA),col=c("red","blue"), lty=c(NA,1) )
```

- By default, R interprets math operators to be element-wise operators. If you want matrix multiplication, you need to add the "%" notation before and after the operator to specify this to R. For example, `A %*% B` does a matrix multiply, while `A*B` does an element-wise multiplication.

### 2.4 Visualizing $J(\theta)$

To understand the cost function $J(\theta)$ better, you will now plot the cost over
a 2-dimensional grid of $\theta_0$ and $\theta_1$ values. You will not need to code anything
new for this part, but you should understand how the code you have written
already is creating these images. In the next step of `ex1.r`, there is code set up to calculate $J(\theta)$ over a grid of values using the `computeCost` function that you wrote.

```{r, echo=F}
# Grid over which we will calculate J
theta0_vals <- seq(-10, 10, length.out=100)
theta1_vals <- seq(-2, 4, length.out=100)
```

```{r}
# initialize J_vals to a matrix of 0's
J_vals <- matrix(0, length(theta0_vals), length(theta1_vals))

# Fill out J_vals
for (i in 1:length(theta0_vals)) {
    for (j in 1:length(theta1_vals)) {
      J_vals[i,j] <- computeCost(X, y, c(theta0_vals[i], theta1_vals[j]))
    }
}
```

After these lines are executed, you will have a 2-D array of $J(\theta)$ values.
The script `ex1.r` will then use these values to produce surface and contour
plots of $J(\theta)$ using the `persp` and `contour` commands. The plots should look
something like Figure 3:

```{r, fig.cap="Figure 3: Cost function $J(\\theta)$", fig.width=14, echo=F}
# dev.new(heigh=7, width=14)
layout(matrix(1:2, 1))
par(mar=rep(1,4))

nbcol = 100
color = rev(rainbow(nbcol, start = 0/6, end = 4/6))
# unlike rgl:persp3d, graphics:persp assigns colors to facets. We compute a facet's z as the sum of its vertices' z.
persp_zfacet <- function (z) {
  ncz <- ncol(z)
  nrz <- nrow(z)
  zfacet <- z[-1, -1] + z[-1, -ncz] + z[-nrz, -1] + z[-nrz, -ncz]
}
J_vals_col  = cut(persp_zfacet(J_vals), nbcol)

pmat <- persp(theta0_vals, theta1_vals, J_vals, col = color[J_vals_col],
        xlab="theta_0", ylab="theta_1",
        zlab="Cost", main = "(a) Gradient Descent", ticktype="detailed", phi=45, theta=45)
t <- trans3d(theta_history[, 1], theta_history[, 2], J_history+10, pmat)
# points(t, col="red")
# lines(t, col="red")

par(mar=c(4,4,3,2))
logspace <- function(d1, d2, n) return(exp(log(10)*seq(d1, d2, length.out=n)))
contour(theta0_vals, theta1_vals, J_vals, levels = logspace(-2, 3, 20), 
        main="(b) Contour showing minimum",
        xlab=expression(theta[0]),
        ylab=expression(theta[1]),
        drawlabels = FALSE)
points(theta[1], theta[2], pch=4, cex=2,col="red",lwd=2)
```

The purpose of these graphs is to show you that how $J(\theta)$ varies with
changes in $\theta_0$ and $\theta_1$. The cost function $J(\theta)$ is bowl-shaped and has a global
mininum. (This is easier to see in the contour plot than in the 3D surface
plot). This minimum is the optimal point for $\theta_0$ and $\theta_1$, and each step of
gradient descent moves closer to this point.

___
#### Optional Exercises

If you have successfully completed the material above, congratulations! You
now understand linear regression and should able to start using it on your
own datasets.
For the rest of this programming exercise, we have included the following
optional exercises. These exercises will help you gain a deeper understanding
of the material, and if you are able to do so, we encourage you to complete
them as well.

___

## 3 Linear regression with multiple variables

In this part, you will implement linear regression with multiple variables to
predict the prices of houses. Suppose you are selling your house and you
want to know what a good market price would be. One way to do this is to
first collect information on recent houses sold and make a model of housing
prices.
The file `ex1data2.txt` contains a training set of housing prices in Portland,
Oregon. The first column is the size of the house (in square feet), the
second column is the number of bedrooms, and the third column is the price
of the house.
The `ex1_multi.r` script has been set up to help you step through this
exercise.

### 3.1 Feature Normalization

The `ex1_multi.r` script will start by loading and displaying some values
from this dataset. By looking at the values, note that house sizes are about
1000 times the number of bedrooms. When features differ by orders of magnitude, first performing feature scaling can make gradient descent converge
much more quickly.

Your task here is to complete the code in `featureNormalize.r` to

- Subtract the mean value of each feature from the dataset.
- After subtracting the mean, additionally scale (divide) the feature values by their respective "standard deviations."

The standard deviation is a way of measuring how much variation there is
in the range of values of a particular feature (most data points will lie within
$\pm 2$ standard deviations of the mean); this is an alternative to taking the range
of values (max-min). In R, you can use the `sd` function to
compute the standard deviation. For example, inside `featureNormalize.r`,
the quantity `X[,1]` contains all the values of $x_1$ (house sizes) in the training
set, so `sd(X[,1])` computes the standard deviation of the house sizes.
At the time that `featureNormalize.r` is called, the extra column of 1's
corresponding to $x_{\theta}=1$ has not yet been added to X (see `ex1_multi.r` for details).
You will do this for all the features and your code should work with
datasets of all sizes (any number of features / examples). Note that each
column of the matrix X corresponds to one feature.

*You should now submit your solutions.*

___
**Implementation Note:** When normalizing the features, it is important
to store the values used for normalization - the mean value and the standard deviation used for the computations. After learning the parameters
from the model, we often want to predict the prices of houses we have not
seen before. Given a new x value (living room area and number of bedrooms), we must first normalize x using the mean and standard deviation
that we had previously computed from the training set.

___
### 3.2 Gradient Descent

Previously, you implemented gradient descent on a univariate regression
problem. The only difference now is that there is one more feature in the
matrix X. The hypothesis function and the batch gradient descent update
rule remain unchanged.
You should complete the code in `computeCostMulti.r` and `gradientDescentMulti.r`
to implement the cost function and gradient descent for linear regression with
multiple variables. If your code in the previous part (single variable) already
supports multiple variables, you can use it here too.
Make sure your code supports any number of features and is well-vectorized.
You can use `nrow(X)` to find out how many rows (objects) are present in the
dataset.

*You should now submit your solutions.*

___
**Implementation Note:** In the multivariate case, the cost function can
also be written in the following vectorized form:

$$J(\theta)=\frac{1}{2m} (X\theta-\vec{y})^T \,(X\theta-\vec{y})$$

where 

$$X=\begin{pmatrix} 
\ldots & (x^{(1)})^T & \ldots \\
\ldots & (x^{(2)})^T & \ldots \\
 & \vdots & \\
\ldots & (x^{(m)})^T & \ldots
\end{pmatrix}
\qquad
\vec{y}=\begin{pmatrix}
y^{(1)} \\
y^{(2)} \\
\vdots \\
y^{(m)}
\end{pmatrix}
$$

The vectorized version is efficient when you're working with numerical
computing tools like R. If you are an expert with matrix
operations, you can prove to yourself that the two forms are equivalent.

___

#### 3.2.1 Optional (ungraded) exercise: Selecting learning rates

In this part of the exercise, you will get to try out different learning rates for
the dataset and find a learning rate that converges quickly. You can change
the learning rate by modifying `ex1_multi.r` and changing the part of the
code that sets the learning rate.
The next phase in `ex1_multi.r` will call your `gradientDescent.r` function and run gradient descent for about 50 iterations at the chosen learning rate. The function should also return the history of $J(\theta)$ values in a vector J. After the last iteration, the `ex1_multi.r` script plots the J values against the number of the iterations.
If you picked a learning rate within a good range, your plot looks similar to
Figure 4. If your graph looks very different, especially if your value of $J(\theta)$
increases or even blows up, adjust your learning rate and try again. We recommend trying values of the learning rate $\alpha$ on a log-scale, at multiplicative
steps of about 3 times the previous value (i.e., 0.3, 0.1, 0.03, 0.01 and so on).
You may also want to adjust the number of iterations you are running if that
will help you see the overall trend in the curve.

```{r, fig.cap="Figure 4: Convergence of gradient descent with an appropriate learning rate", echo=F}
data <- read.table('ex1data2.txt',sep = ',')
X <- data[,1:2]
y <- data[,3]
m <- length(y)
fN <- featureNormalize(X)
X <- fN$X_norm
mu <- fN$mu
sigma <- fN$sigma

# Add intercept term to X
X <- cbind(rep(1,m),X)
X <- as.matrix(X)

# Choose some alpha value
alpha <- 1 # modified from 0.01 because 3.2.1
num_iters <- 50 #modified from 100 because 3.2.1

# Init Theta and Run Gradient Descent
theta <- rep(0,3)
# Here we can test different learning parameter alpha
gDM <- gradientDescentMulti(X, y, theta, alpha , num_iters)
theta <- gDM$theta
J_history <- gDM$J_history
rm(gDM)

# Plot the convergence graph
par(mar=c(3,3,.5,0), mgp=c(1.5,.5,0))
plot(1:length(J_history), J_history, type="l", col="blue", lwd=2, cex=.1,
     xlab="Number of Iterations", ylab="Cost J")
```

___
**Implementation Note:** If your learning rate is too large, $J(\theta)$ can diverge and 'blow up', resulting in values which are too large for computer
calculations. In these situations, R will tend to return
`NaNs` or `Inf`. NaN stands for 'not a number' and is often caused by undefined operations that involve $\pm\infty$. 

___
**R Tip:** To compare how different learning learning
rates affect convergence, it's helpful to plot J for several learning rates
on the same figure. In R, this can be done by first setting up the first `plot(..., type='l')` and then calling `lines(...)` multiple times. Concretely, if you've tried three different values of $\alpha$ (you should
probably try more values than this) and stored the costs in J1, J2 and
J3, you can use the following commands to plot them on the same figure:

```{r, eval=FALSE}
plot(1:50, J1[1:50], col=1, type='l')
lines(1:50, J2[1:50], col=2)
lines(1:50, J3[1:50], col=3)
```

The final argument `col` specifies different colors for the plots.

___
Notice the changes in the convergence curves as the learning rate changes.
With a small learning rate, you should find that gradient descent takes a very
long time to converge to the optimal value. Conversely, with a large learning
rate, gradient descent might not converge or might even diverge!
Using the best learning rate that you found, run the `ex1_multi.r` script
to run gradient descent until convergence to find the final values of $\theta$. Next,
use this value of $\theta$ to predict the price of a house with 1650 square feet and
3 bedrooms. You will use this value later to check your implementation of the
normal equations. Don't forget to normalize your features when you make
this prediction!

You do not need to submit any solutions for these optional (ungraded)
exercises.

### 3.3 Normal Equations

In the lecture videos, you learned that the closed-form solution to linear
regression is

$$ \theta = (X^T X)^{-1} X^T \vec{y}. $$

Using this formula does not require any feature scaling, and you will get
an exact solution in one calculation: there is no "loop until convergence" like in gradient descent.
Complete the code in `normalEqn.r` to use the formula above to calculate $\theta$. Remember that while you don't need to scale your features, we still
need to add a column of 1's to the X matrix to have an intercept term ($\theta_0$).
The code in `ex1.r` will add the column of 1's to X for you.

*You should now submit your solutions.*

*Optional (ungraded) exercise:* Now, once you have found $\theta$ using this
method, use it to make a price prediction for a 1650-square-foot house with
3 bedrooms. You should find that gives the same predicted price as the value
you obtained using the model fit with gradient descent (in Section 3.2.1).

## Submission and Grading
After completing various parts of the assignment, be sure to use the submit
function to submit your solutions to our servers. The following is a
breakdown of how each part of this exercise is scored.

| Part                              | Submitted File    | Points     |
| :----------------                              | :----:            | :--------: |
| Warm up exercise                  | `warmUpExercise.r`  | 10 points  |
| Compute cost for one variable     | `computeCost.r`     | 40 points  |
| Gradient descent for one variable | `gradientDescent.r` | 50 points  |
| Total Points                      |                   | 100 points |

### Optional Exercises


| Part                                    | Submitted File           | Points     |
| :-----------------                                | :----:                   | :--------: |
| Feature normalization                   | `featureNormalize.r`     | 0 points   |
| Compute cost for multiple variables     | `computeCostMulti.r`     | 0 points   |
| Gradient descent for multiple variables | `gradientDescentMulti.r` | 0 points   |
| Normal Equations                        | `normalEqn.r`            | 0 points   |

You are allowed to submit your solutions multiple times, and we will take
only the highest score into consideration.
