---
title: "Programming Exercise 2: Logistic Regression"
author: ""
date: "Dec 21, 2020"
output: 
    html_document:
        toc: true
        mathjax: local
        self_contained: FALSE
---

```{r opt, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache=TRUE, fig.align="center")
sources <- c("costFunction.r","sigmoid.r","plotData.r",
             "plotDecisionBoundary.r","mapFeature.r",
             "predict.r","costFunctionReg.r")
sources <- paste0(substr(sources, 1, nchar(sources)-2), '-solution.r')
invisible(lapply(sources, source))
```

This programming exercise instruction was originally developed and written by Prof. Andrew
Ng as part of his machine learning [course](https://www.coursera.org/learn/machine-learning) on Coursera platform. I have adapted
the instruction for R language, so that its users, including myself, could also take and benefit from the course.

## Introduction

In this exercise, you will implement logistic regression and apply it to two
different datasets. Before starting on the programming exercise, we strongly
recommend watching the video lectures and completing the review questions
for the associated topics.
To get started with the exercise, you will need to download the starter
code and unzip its contents to the directory where you wish to complete the
exercise. If needed, use the `setwd()` command in R to change to this directory before starting this exercise.

Files included in this exercise:\

- `ex2.r` - R script that steps you through the exercise
- `ex2_reg.r` - R script for the later parts of the exercise
- `ex2data1.txt` - Training set for the first half of the exercise
- `ex2data2.txt` - Training set for the second half of the exercise
- `submit.r` - Submission script that sends your solutions to our servers
- `mapFeature.r` - Function to generate polynomial features
- `plotDecisionBoundary.r` - Function to plot classifier's decision boundary
- ⋆ `plotData.r` - Function to plot 2D classification data
- ⋆ `sigmoid.r` - Sigmoid Function
- ⋆ `costFunction.r` - Logistic Regression Cost Function
- ⋆ `predict.r` - Logistic Regression Prediction Function
- ⋆ `costFunctionReg.r` - Regularized Logistic Regression Cost

⋆ indicates files you will need to complete

Throughout the exercise, you will be using the scripts `ex2.r` and `ex2_reg.r`.
These scripts set up the dataset for the problems and make calls to functions
that you will write. You do not need to modify either of them. You are only
required to modify functions in other files, by following the instructions in
this assignment.

### Where to get help
The exercises in this course use R, a high-level programming language
well-suited for numerical computations. If you do not have R installed, please
download a Windows installer from [R-project](https://cran.r-project.org/bin/windows/base/) website. [R-Studio](https://rstudio.com/products/rstudio/download/) is a free and open-source R integrated development environment (IDE) making R script development a bit easier when compared to the R's own basic GUI. You may start from the `.rproj` (a R-Studio project file) in each exercise directory. 
At the R command line, typing help followed by a
function name displays documentation for that function. For example,
`help('plot')` or simply `?plot` will bring up help information for plotting.
Further documentation for R functions can be found at the R documentation pages. 

## 1 Logistic Regression
In this part of the exercise, you will build a logistic regression model to
predict whether a student gets admitted into a university.
Suppose that you are the administrator of a university department and
you want to determine each applicant's chance of admission based on their
results on two exams.
You have historical data from previous applicants
that you can use as a training set for logistic regression. For each training
example, you have the applicant's scores on two exams and the admissions
decision.
Your task is to build a classification model that estimates an applicant's
probability of admission based on the scores from those two exams. This outline
and the framework code in `ex2.r` will guide you through the exercise.

### 1.1 Visualizing the data
Before starting to implement any learning algorithm, it is always good to
visualize the data if possible. In the first part of `ex2.r`, the code will load the
data and display it on a 2-dimensional plot by calling the function `plotData`.
You will now complete the code in `plotData` so that it displays a figure
like Figure 1, where the axes are the two exam scores, and the positive and
negative examples are shown with different markers.

```{r , fig.cap='Figure 1: Scatter plot of training data', echo=FALSE}
par(mar=c(4,4,1,1))
data <- read.table('ex2data1.txt', sep = ',')
X <- data[,c(1,2)]; y <- data[,3]
X <- as.matrix(X)
plotData(X, y)
```

To help you get more familiar with plotting, we have left `plotData.r`
empty so you can try to implement it yourself. However, this is an optional
(ungraded) exercise. We also provide our implementation below so you can
copy it or refer to it. If you choose to copy our example, make sure you learn
what each of its commands is doing by consulting the R documentation.

```{r, eval=F}
# plus and empty circle character codes
symbols <- c(3,21) 
yfac <- factor(y)
plot(X[,1], X[,2], pch = symbols[yfac], 
     bg = "yellow", lwd = 1.3, xlab = axLables[1], ylab = axLables[2])
legend("topright", legLabels, pch = rev(symbols), pt.bg = "yellow")
```

### 1.2 Implementation

#### 1.2.1 Warmup exercise: sigmoid function
Before you start with the actual cost function, recall that the logistic regression hypothesis is defined as:

$$ h_\theta(x)=g(\theta^Tx),$$

where function $g$ is the sigmoid function. The sigmoid function is defined as:

$$ g(z)=\frac{1}{1+e^{-z}}.$$

Your first step is to implement this function in `sigmoid.r` so it can be
called by the rest of your program. When you are finished, try testing a few
values by calling `sigmoid(x)` at the R command line. For
large positive values of x, the sigmoid should be close to 1, while for large
negative values, the sigmoid should be close to 0. Evaluating `sigmoid(0)`
should give you exactly 0.5. Your code should also work with vectors and
matrices. For a matrix, your function should perform the sigmoid function on every element.
You can submit your solution for grading by typing `submit()` at the R command line. The submission script will prompt you for
your login e-mail and submission token and ask you which files you want
to submit. You can obtain a submission token from the web page for the assignment.

*You should now submit your solutions.*

### 1.2.2 Cost function and gradient
Now you will implement the cost function and gradient for logistic regression.
Complete the code in `costFunction.r` to return the cost and gradient.
Recall that the cost function in logistic regression is

$$J(\theta)=\frac{1}{m}\sum_{i=1}^m [-y^{(i)}log(h_\theta(x^{(i)})) - (1-y^{(i)}) log(1-h_\theta(x^{(i)})) ],$$

and the gradient of the cost is a vector of the same length as $\theta$ where
the $j^{th}$ element (for $j = 0, 1, . . . , n$) is defined as follows:

$$ \frac{\partial J(\theta)}{\partial \theta_j} = \frac{1}{m}\sum_{i=1}^m (h_\theta(x^{(i)}) - y^{(i)}) x_j^{(i)} $$

Note that while this gradient looks identical to the linear regression
gradient, the formula is actually different because linear and logistic
regression have different definitions of $h_\theta(x)$.
Once you are done, `ex2.r` will call your `costFunction` using the initial
parameters of $\theta$. You should see that the cost is about 0.693.

*You should now submit your solutions.*

#### 1.2.3 Learning parameters using optim [^0]
In the previous assignment, you found the optimal parameters of a linear regression model by implementing gradent descent. You wrote a cost function
and calculated its gradient, then took a gradient descent step accordingly.
This time, instead of taking gradient descent steps, you will use an R built-in function called `optim`.
R's `optim` is an optimization solver that finds the minimum of an unconstrained [^1] function. For logistic regression, you want to
optimize the cost function $J(\theta)$ with parameters $\theta$.
Concretely, you are going to use `optim` to find the best parameters $\theta$
for the logistic regression cost function, given a fixed dataset (of X and y values). You will pass to `optim` the following inputs:
 
[^0]: Beware that `fmincg` or `fminunc` optimization function in MATLAB takes one function as input and computes cost and gradient simultaneously. However, cost and gradient functions must be supplied into `optim` or `lbfgsb3` functions individually.

[^1]: Constraints in optimization often refer to constraints on the parameters, for example,
constraints that bound the possible values $\theta$ can take (e.g., $\theta\leq 1$). Logistic regression
does not have such constraints since $\theta$ is allowed to take any real value.

- The initial values of the parameters we are trying to optimize.
- Two functions that, when given the training set and a particular $\theta$, computes the logistic regression cost and gradient respectively with respect to $\theta$ for the dataset (X, y)

In `ex2.r`, we already have code written to call `optim` with the correct arguments.

```{r, eval=F}
optimRes <- optim(par = initial_theta, fn = costFunction(X,y), gr = grad(X,y), 
      method="BFGS", control = list(maxit = 400))
theta <- optimRes$par
cost <- optimRes$value
```

In this code snippet, we first defined the options to be used with `optim`.
Specifically, we set the `gr` parameter to a function to tell `optim` to use the gradient when minimizing the function. If it is unset, `optim` uses a finite difference approximation to compute gradient.
Method BFGS is a quasi-Newton method (also known as a variable metric
algorithm). This uses both the function values and gradients to build up a
picture of the surface to be optimized.
Furthermore, we set the `maxit` option to 400, so that `optim` will run for at most 400 steps before
it terminates.

If you have completed the `costFunction` correctly, `optim` will converge
on the right optimization parameters and return the final values of the cost
and $\theta$. Notice that by using `optim`, you did not have to write any loops
yourself, or set a learning rate like you did for gradient descent. This is all
done by `optim`: you only needed to provide a function calculating the cost
and another function for the gradient.

Once `optim` completes, `ex2.r` will call your `costFunction` function
using the optimal parameters of $\theta$. You should see that the cost is about 0.203.
This final $\theta$ value will then be used to plot the decision boundary on the
training data, resulting in a figure similar to Figure 2. We also encourage
you to look at the code in `plotDecisionBoundary.r` to see how to plot such
a boundary using the $\theta$ values.

#### 1.2.4 Evaluating logistic regression
After learning the parameters, you can use the model to predict whether a
particular student will be admitted. For a student with an Exam 1 score
of 45 and an Exam 2 score of 85, you should expect to see an admission
probability of 0.776.
Another way to evaluate the quality of the parameters we have found
is to see how well the learned model predicts on our training set. 

```{r, fig.cap="Figure 2: Training data with decision boundary", echo=FALSE}
par(mar=c(4,4,1,1))
#  Setup the data matrix appropriately, and add ones for the intercept term
m <- dim(X)[1]
n <- dim(X)[2]

# Add intercept term to x and X_test
X <- cbind(rep(1,m),X)
# Initialize fitting parameters
initial_theta <- rep(0,n+1)

# Compute and display initial cost and gradient
cF <- costFunction(X, y)(initial_theta)
cost <- costFunction(X, y)(initial_theta)
grd <- grad(X,y)(initial_theta)

optimRes <- optim(par = initial_theta, fn = costFunction(X,y), gr = grad(X,y), 
      method="BFGS", control = list(maxit = 400))
theta <- optimRes$par
cost <- optimRes$value

# Plot Boundary
plotDecisionBoundary(theta, X, y)
```

In this part, your task is to complete the code in `predict.r`. The `predict` function
will produce 1 or 0 predictions given a dataset and a learned parameter
vector $\theta$.
After you have completed the code in `predict.r`, the `ex2.r` script will
proceed to report the training accuracy of your classifier by computing the
percentage of examples it got correct.

*You should now submit your solutions.*

## 2 Regularized logistic regression
In this part of the exercise, you will implement regularized logistic regression
to predict whether microchips from a fabrication plant passes quality assurance (QA). 
During QA, each microchip goes through various tests to ensure it is functioning correctly.
Suppose you are the product manager of the factory and you have the
test results for some microchips on two different tests. From these two tests,
you would like to determine whether the microchips should be accepted or
rejected. To help you make the decision, you have a dataset of test results
on past microchips, from which you can build a logistic regression model.

You will use another script, `ex2_reg.r` to complete this portion of the exercise.

### 2.1 Visualizing the data
Similar to the previous parts of this exercise, `plotData` is used to generate a
figure like Figure 3, where the axes are the two test scores, and the positive
(y = 1, accepted) and negative (y = 0, rejected) examples are shown with
different markers.

```{r, fig.cap="Figure 3: Plot of training data", echo=F}
par(mar=c(4,4,1,1))
data <- read.table('ex2data2.txt', sep = ',')
X <- data[,c(1,2)]; y <- data[,3]
X <- as.matrix(X)

plotData(X, y, axLables = c('Microchip Test 1','Microchip Test 2'),
         legLabels = c("y = 1","y = 0"))
```

Figure 3 shows that our dataset cannot be separated into positive and
negative examples by a straight-line through the plot. Therefore, a
straightforward application of logistic regression will not perform well on
this dataset since logistic regression will only be able to find a linear
decision boundary.

### 2.2 Feature mapping

One way to fit the data better is to create more features from each data
point. In the provided function `mapFeature.r`, we will map the features into
all polynomial terms of $x_1$ and $x_2$ up to the sixth power.

$$ 
mapFeature(x) = \left[
\begin{array}{c}
1 \\
x_1 \\
x_2 \\
x_1^2 \\
x_1x_2 \\
x_2^2 \\
x_1^3 \\
\vdots \\
x_1x_2^5 \\
x_2^6
\end{array}
\right]
$$

As a result of this mapping, our vector of two features (the scores on two QA
tests) has been transformed into a 28-dimensional vector. A logistic regression
classifier trained on this higher-dimension feature vector will have
a more complex decision boundary and will appear nonlinear when drawn in
our 2-dimensional plot.
While the feature mapping allows us to build a more expressive classifier,
it is also more susceptible to overfitting. In the next parts of the exercise, you
will implement regularized logistic regression to fit the data and also see for
yourself how regularization can help combat the overfitting problem.

### 2.3 Cost function and gradient

Now you will implement code to compute the cost function and gradient for
regularized logistic regression. Complete the code in `costFunctionReg.r` to
return the cost and gradient. Recall that the regularized cost function in logistic regression is

$$
J(\theta)=\frac{1}{m}\sum_{i=1}^m [-y^{(i)} log(h_\theta(x^{(i)})) - 
(1-y^{(i)}) log(1-h_\theta(x^{(i)})) ] +
\frac{\lambda}{2m} \sum_{j=1}^n \theta_j^2.
$$

Note that you should not regularize the parameter $\theta_0$. In R, recall that
indexing starts from 1, hence, you should not be regularizing the `theta[1]`
parameter (which corresponds to $\theta_0$) in the code. The gradient
of the cost function is a vector where the $j^{th}$ element is defined as follows:

<center>
$\displaystyle \frac{\partial J(\theta)}{\partial \theta_0} = \frac{1}{m}\sum_{i=1}^m (h_\theta(x^{(i)}) - y^{(i)}) x_j^{(i)} \qquad$  for $j=0$

$\displaystyle \frac{\partial J(\theta)}{\partial \theta_j} = \left (\frac{1}{m}\sum_{i=1}^m (h_\theta(x^{(i)}) - y^{(i)}) x_j^{(i)} \right) + \frac{\lambda}{m}\theta_j \qquad$ for $j \geq 1$
</center>

Once you are done, `ex2_reg.r` will call your `costFunctionReg` function using the initial value of $\theta$ (initialized to all zeros). You should see that the cost is about 0.693.

*You should now submit your solutions.*

#### 2.3.1 Learning parameters using `optim`

Similar to the previous parts, you will use `optim` to learn the optimal
parameters $\theta$. If you have completed the cost and gradient for regularized
logistic regression (`costFunctionReg.r`) correctly, you should be able to step
through the next part of `ex2_reg.r` to learn the parameters $\theta$ using `optim`.

### 2.4 Plotting the decision boundary

To help you visualize the model learned by this classifier, we have provided 
the function `plotDecisionBoundary.r` which plots the (non-linear)
decision boundary that separates the positive and negative examples.
In `plotDecisionBoundary.r`, we plot the non-linear decision boundary by 
computing the classifier's predictions on an evenly spaced grid and then drew
a contour plot of where the predictions change from $y = 0$ to $y = 1$.
After learning the parameters $\theta$, the next step in the `ex2_reg.r` will plot a
decision boundary similar to Figure 4.

### 2.5 Optional (ungraded) exercises

In this part of the exercise, you will get to try out different regularization
parameters for the dataset to understand how regularization prevents overfitting.
Notice the changes in the decision boundary as you vary $\lambda$. With a small
$\lambda$, you should find that the classifier gets almost every training example
correct, but draws a very complicated boundary, thus overfitting the data
(Figure 5). This is not a good decision boundary: for example, it predicts
that a point at x = (−0.25, 1.5) is accepted (y = 1), which seems to be an
incorrect decision given the training set.
With a larger $\lambda$, you should see a plot that shows a simpler decision
boundary which still separates the positives and negatives fairly well. However,
if $\lambda$ is set to too high a value, you will not get a good fit and the decision
boundary will not follow the data so well, thus underfitting the data (Figure 6).

You do not need to submit any solutions for these optional (ungraded) exercises.

```{r, fig.cap="Figure 4: Training data with decision boundary ($\lambda = 1$)", echo=FALSE}
par(mar=c(4,4,1,1))
X <- mapFeature(X[,1], X[,2])

initial_theta <- rep(0,dim(X)[2])
lambda <- 1
optimRes <- optim(par = initial_theta, 
                  fn = costFunctionReg(X,y,lambda), 
                  gr = gradReg(X,y,lambda), 
                  method="BFGS", 
                  control = list(maxit = 400))

theta <- optimRes$par
J <- optimRes$value
plotDecisionBoundary(theta, X, y, axLables = c('Microchip Test 1','Microchip Test 2'),
                     legLabels = c("y = 1","y = 0"))
```

```{r, fig.cap="Figure 5: No regularization (Overfitting) ($\lambda= 0$)", echo=FALSE}
par(mar=c(4,4,1,1))
lambda <- 0
optimRes <- optim(par = initial_theta, 
                  fn = costFunctionReg(X,y,lambda), 
                  gr = gradReg(X,y,lambda), 
                  method="BFGS", 
                  control = list(maxit = 400))

theta <- optimRes$par
J <- optimRes$value
plotDecisionBoundary(theta, X, y, axLables = c('Microchip Test 1','Microchip Test 2'),
                     legLabels = c("y = 1","y = 0"))
```

```{r, fig.cap="Figure 6: Too much regularization (Underfitting) ($\lambda=100$)", echo=FALSE}
par(mar=c(4,4,1,1))
lambda <- 100
optimRes <- optim(par = initial_theta, 
                  fn = costFunctionReg(X,y,lambda), 
                  gr = gradReg(X,y,lambda), 
                  method="BFGS", 
                  control = list(maxit = 400))

theta <- optimRes$par
J <- optimRes$value
plotDecisionBoundary(theta, X, y, axLables = c('Microchip Test 1','Microchip Test 2'),
                     legLabels = c("y = 1","y = 0"))
```

## Submission and Grading

After completing various parts of the assignment, be sure to use the submit
function system to submit your solutions to our servers. The following is a
breakdown of how each part of this exercise is scored.

| Part                                 | Submitted File      | Points     |
| :-------------                              | :----                     | :--------: |
| Sigmoid Function                     | `sigmoid.r`         | 5 points   |
| Compute cost for logistic regression | `costFunction.r`    | 30 points  |
| Gradient for logistic regression     | `costFunction.r`    | 30 points  |
| Predict Function                     | `predict.r`         | 5 points   |
| Compute cost for regularized LR      | `costFunctionReg.r` | 15 points  |
| Gradient for regularized LR          | `costFunctionReg.r` | 15 points  |
| Total Points                         |                     | 100 points |

You are allowed to submit your solutions multiple times, and we will take only the highest score into consideration.
